---
title: "Climatekit: Unified Climate Indices for Temperature, Precipitation, and Drought in R"
date: "2026-04-19"
abstract: >
  The climatekit package computes the standard suite of climate
  indices from daily weather observations behind a single R
  interface. Thirty functions cover temperature (frost days, ice
  days, summer days, growing and heating degree days, tropical
  nights, diurnal range, growing-season length, warm-spell duration),
  precipitation (wet and dry spells, heavy-precipitation days,
  one-day and five-day maxima, precipitation intensity), drought
  (Standardised Precipitation Index, Standardised
  Precipitation-Evapotranspiration Index, potential
  evapotranspiration via the Hargreaves method), agroclimatology
  (Huglin, Winkler, Branas, frost dates), and thermal comfort (wind
  chill, heat index, humidex, fire danger). Definitions follow the
  Expert Team on Sector-specific Climate Indices recommendations
  where applicable. Every function returns a tidy data frame with a
  period, a value, an index name, and a unit, so outputs compose
  directly. The package is pure computation, has no API calls, and
  is available on CRAN.
draft: false
author:
  - name: Charles Coverdale
    affiliation: Independent
    address:
    - London, United Kingdom
    url: https://github.com/charlescoverdale/climatekit
    email: charles.f.coverdale@gmail.com
type: package
journal:
  section: "Research article"
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
header-includes:
  - \renewcommand{\address}[1]{}
bibliography: RJreferences.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", out.width = "100%")
```

# Introduction

The \CRANpkg{climatekit} package consolidates thirty climate
indices behind a single consistent interface. Each function takes
daily vectors (temperature, precipitation, humidity) with an
accompanying date vector and returns a data frame with four
columns: `period` (the reporting period, typically annual or
monthly), `value` (the index value), `index` (a character name),
and `unit`. Users can feed any output directly into plotting,
tabulation, or further aggregation without reshaping.

Climate-index analysis is a routine operation across applied
climate work: national meteorological services publish annual
index tables, regional climate assessments benchmark stations
against one another, agricultural and viticultural research tracks
growing-season indices, insurance underwriters use drought indices
in parametric products, and public-health authorities monitor
heat-index exceedances. Until now, the R ecosystem has exposed
parts of this analysis piecemeal, with separate packages for
drought indices (\CRANpkg{SPEI}), for standard climate-change
indices (\pkg{climdex.pcic}, archived), and for various
agricultural indices. \CRANpkg{climatekit} unifies the set.

The package has no network access, no external data dependency,
and no runtime requirements beyond \CRANpkg{cli} and base R. Users
pair it with a data-acquisition package (\CRANpkg{readnoaa} for
NOAA's Global Historical Climatology Network, or direct downloads
from a national met service) and pass the resulting daily series
straight in.

# Background

Climate-index standardisation dates to @frich2002observed, who
proposed twenty-seven indices targeted at climate-change
detection, and continued through the joint CCl/WCRP/JCOMM Expert
Team on Climate Change Detection and Indices (ETCCDI), now the
Expert Team on Sector-specific Climate Indices (ET-SCI).
@zhang2011indices synthesised the ET-SCI core set, which informs
most modern national climate assessments. Applied analyses of the
ETCCDI indices on observed station records [@alexander2006global]
and on gridded reconstructions [@donat2013hadex2] document robust
multi-decadal trends. @sillmann2013indices evaluate the same
indices across CMIP5 climate-model output, establishing the index
suite as the bridge between observational and model-based
climate-change attribution.

**Temperature indices.** The core definitions are threshold counts
(frost days where $T_{\min} < 0^\circ$C; tropical nights where
$T_{\min} \geq 20$; summer days where $T_{\max} \geq 25$; ice days
where $T_{\max} < 0$) and degree-day accumulations (growing degree
days above a base, heating degree days below a reference, cooling
degree days above a reference). Diurnal range, growing-season
length, and warm-spell duration round out the temperature family.

**Precipitation indices.** Consecutive dry and wet days summarise
spell length. Total accumulation, heavy-precipitation exceedances,
and one- and five-day maxima summarise intensity. The simple daily
intensity index (SDII) is the average daily total on wet days. The
ETCCDI-standard thresholds for heavy-precipitation counts are the
95th (R95p) and 99th (R99p) percentiles of the wet-day
distribution at each station; `ck_heavy_precip()` accepts an
explicit threshold argument so users can supply either a fixed
value or a station-specific percentile computed from the data.

**Drought indices.** @mckee1993relationship introduced the
Standardised Precipitation Index (SPI), which fits a gamma
distribution to precipitation accumulated over a chosen window
(three, six, twelve months) and standardises to a standard normal.
Values below minus one indicate moderate drought; below minus
two, extreme drought. @vicente2010multiscalar extended this to
the Standardised Precipitation-Evapotranspiration Index (SPEI),
which incorporates potential evapotranspiration via a climatic
water balance. @stagge2015candidate compare candidate
distributions for both indices and find that the gamma family (for
SPI) and the generalised logistic (for SPEI) perform best in
normality-of-residuals tests; \CRANpkg{climatekit} uses these
defaults. Potential evapotranspiration is computed by the
@hargreaves1985reference method from temperature extremes and
extra-terrestrial radiation.

**Agroclimatic indices.** The Huglin heliothermal index
[@huglin1978nouveau] accumulates heat summed between a base
temperature and a latitude-adjusted day length; it is the primary
viticultural suitability index in temperate Europe. The Winkler
index [@winkler1974general] is a similar cumulative heat metric
used widely in California. Branas, Bernon, and Levadoux's index is
the French viticultural analogue. First-frost and last-frost dates
set the frost-free growing window.

**Thermal comfort.** The US National Weather Service's heat index
[@rothfusz1990heat] converts temperature and humidity into an
apparent-temperature equivalent; Canadian humidex
[@masterton1979humidex] does the same with a different functional
form. Wind chill [@osczevski2005new] is the cold-weather
counterpart. Fire-weather indices combine temperature, humidity,
wind, and precipitation history.

# Package design

\CRANpkg{climatekit} is pure R with no compiled code. Runtime
imports are \CRANpkg{cli}, `stats`, and `tools`. R 4.1.0 or later
is required. The test suite contains over a hundred tests covering
every exported function and every index definition.

Every index function returns a data frame with four columns:
`period`, `value`, `index`, and `unit`. The `period` is a `Date`
keyed to the start of the reporting window (year or month); the
`value` is numeric; `index` and `unit` are character strings. This
invariant means output from any function composes directly into a
larger panel via `rbind()`, and plotting code is shared across
index families. Every index function takes one or two daily
numeric vectors plus a `dates` vector of matching length.

`ck_compute()` dispatches to any index by name, letting users
select the index programmatically from a configuration file or
argument. `ck_available()` lists the indices that can be computed
given a supplied set of inputs; `ck_metadata()` returns each
index's definition, unit, reference, and typical use.

# Temperature indices

Four threshold-exceedance counters cover the core temperature
indices. Cold extremes are counted by `ck_frost_days()` (minimum
below $0^\circ$C) and `ck_ice_days()` (daytime maximum below
zero). Hot extremes are counted by `ck_summer_days()` (maximum at
or above 25) and `ck_tropical_nights()` (minimum at or above 20).
Three degree-day accumulators report cumulative heat above or
below a reference temperature. Agricultural work uses
`ck_growing_degree_days()` to sum degree-days above a
user-specified base. Energy-demand applications use the heating
and cooling counterparts, `ck_heating_degree_days()` and
`ck_cooling_degree_days()`. Figure \@ref(fig:central-park) in the
case study below shows these indices applied to seventy-five years
of real NOAA data from the New York Central Park station.

# Precipitation and drought indices

`ck_total_precip()`, `ck_dry_days()`, and `ck_wet_days()`
summarise accumulation and spell counts. `ck_heavy_precip()` and
`ck_very_heavy_precip()` count exceedances of user-specified
thresholds. `ck_max_1day_precip()` and `ck_max_5day_precip()`
return rolling maxima. `ck_precip_intensity()` returns the simple
daily intensity index.

Figure \@ref(fig:precip) shows two precipitation indices on the
Central Park station's daily record from 1950 to 2024: annual
total precipitation (blue) and the count of heavy-precipitation
days with daily total above 10 mm (red, scaled by 50 for visual
comparability). Both trend upward over the seventy-five-year
window: the annual total rises by about 45 mm per decade and the
heavy-precipitation day count by 1.15 days per decade, consistent
with the expected intensification of the hydrological cycle under
continental-US warming documented in @alexander2006global and
@donat2013hadex2.

```{r precip, fig.cap = "Annual precipitation indices for New York Central Park (GHCND station USW00094728), 1950 to 2024. Blue solid: annual total precipitation in millimetres, linear trend $+45$ mm per decade. Red dashed: count of days with daily total above 10 mm, scaled by 50 for visual comparability, linear trend $+1.15$ days per decade. Dotted lines are OLS trend fits. Both indices rise over the period.", fig.alt = "A line chart showing two time series from 1950 to 2024 for New York Central Park. The blue solid line is annual total precipitation in millimetres, rising from around 1000 millimetres at the start of the period to around 1300 by the end. The red dashed line is the count of days per year with total precipitation above 10 millimetres, also rising over time. Dotted regression lines overlay each series, both sloping upward."}
knitr::include_graphics("figures/fig2_precip.pdf")
```

# Drought and pluvial episodes

`ck_spi()` and `ck_spei()` return the Standardised Precipitation
Index and the Standardised Precipitation-Evapotranspiration Index.
Both are unitless anomalies scaled to standard deviations: a value
of $-1$ sits one standard deviation below the long-run mean for
that calendar month, commonly interpreted as moderate drought, and
values below $-2$ as severe drought. The index is computed by
fitting a gamma distribution to rolling accumulated precipitation
totals and transforming to standard normal deviates, separately
for each calendar month to remove seasonality.

Figure \@ref(fig:spi) applies `ck_spi()` at a twelve-month scale
to the 1950-to-2024 daily precipitation record at New York Central
Park. The 1960s Northeast drought is visible as an extended run
of deeply negative values between 1962 and 1966, the most severe
drought episode in the station's recorded history. The 2001-02
drought is also clearly marked. The upper tail is populated by
well-known wet years, including 1972, 1983-84, and 2011.

```{r spi, fig.cap = "Twelve-month Standardised Precipitation Index for New York Central Park, 1950 to 2024. Computed with \\texttt{ck\\_spi(scale = 12)} on the full GHCND USW00094728 daily precipitation record. Dashed horizontal lines at plus and minus one mark the moderate drought and moderate wet thresholds. The 1960s Northeast drought (1962 to 1966) and 2001 to 2002 drought are the two deepest dry episodes; 2011 is the wettest year in the sample.", fig.alt = "A column chart of 12-month SPI values at monthly resolution for New York Central Park, 1950 to 2024. Columns below zero are coloured red and represent dry anomalies; columns above zero are coloured blue and represent wet anomalies. A sustained run of red columns between 1962 and 1966 reaches values around minus 2.3, labelled as the 1960s Northeast drought. Another deep red cluster around 2001 to 2002 is labelled. The deepest blue column, around 2011, reaches about plus 2.9 and is labelled as a wet year."}
knitr::include_graphics("figures/fig5_heat.pdf")
```

# Agroclimatic and comfort indices

`ck_huglin()` returns the Huglin heliothermal index, accumulated
from April through September (Northern Hemisphere) from daily
maximum and mean temperatures weighted by a latitude correction.
`ck_winkler()` returns the Winkler index. `ck_branas()` returns
the Branas product index. `ck_first_frost()` and `ck_last_frost()`
report frost dates each year. Figure \@ref(fig:huglin) shows the
Huglin index applied to ten years of a synthetic temperate site.
Grey horizontal bands mark approximate viticultural classification
ranges; the site plots within the Cool to Temperate range.

```{r huglin, fig.cap = "Huglin heliothermal index on a 10-year synthetic temperate site at latitude 51.5 degrees north. Shaded horizontal bands mark approximate viticultural suitability ranges: below 1500 is too cool, 1500 to 1800 is Cool, 1800 to 2100 is Temperate, 2100 to 2400 is Warm, above 2400 is Hot. The sample falls mostly in the Cool range.", fig.alt = "A line chart of the Huglin heliothermal index across 10 years of a synthetic temperate site. Values fluctuate between about 1550 and 1620, sitting inside a shaded horizontal band labelled Cool. Bands above and below mark Temperate, Warm, Hot, and Too cool ranges."}
knitr::include_graphics("figures/fig4_huglin.pdf")
```

Four comfort functions cover hot and cold extremes.
`ck_heat_index()` returns the US National Weather Service heat
index (apparent temperature) from mean temperature and relative
humidity. `ck_humidex()` returns the Canadian humidex from the
same inputs. `ck_wind_chill()` returns wind chill from temperature
and wind speed. `ck_fire_danger()` returns a linearised
approximation of the Canadian Fire Weather Index (CFFWIS). Users
requiring the full iterative CFFWIS should pair with the
specialised \CRANpkg{cffdrs} package.

# A seventy-five-year case study

Figure \@ref(fig:central-park) applies \CRANpkg{climatekit} to
real NOAA Global Historical Climatology Network Daily (GHCND) data
from the New York Central Park station (USW00094728) from 1950 to
2024, a seventy-five-year daily record with zero missing
observations in $T_{\min}$, $T_{\max}$, and precipitation. Three
temperature indices are computed directly from the raw daily
vectors: frost days, summer days, and growing degree days at base
10. A linear trend fitted over the sample returns a decline of
2.6 frost days per decade, a rise of 28 growing-degree-days per
decade, and a roughly stable summer-days count. The signs and
magnitudes are consistent with the observed warming trends for the
US Northeast documented in @alexander2006global and
@donat2013hadex2. The full computation is three function calls on
three vectors.

```{r central-park, fig.cap = "Three temperature indices for New York Central Park (GHCND station USW00094728), 1950 to 2024. Annual counts from raw daily $T_{\\min}$, $T_{\\max}$, and $T_{\\mathrm{mean}}$. Blue solid: frost days ($T_{\\min} < 0^\\circ$C), linear trend $-2.6$ days per decade. Red dashed: summer days ($T_{\\max} \\geq 25^\\circ$C), trend $+0.3$ days per decade. Green dotdash: growing degree days above base $10^\\circ$C, divided by 20 for visual scale, trend $+28$ GDD per decade. Dotted lines are OLS trend fits.", fig.alt = "A line chart showing three annual temperature-index series for New York Central Park, 1950 to 2024. A blue solid line for frost days falls from about 130 per year in the 1950s to about 100 by the 2020s. A red dashed line for summer days is roughly flat around 45 per year. A green dotdash line for growing degree days (divided by 20) rises from about 80 in the 1950s to about 110 by the 2020s. Thin dotted lines show the linear trend fitted to each series."}
knitr::include_graphics("figures/fig6_central_park.pdf")
```

# Replication

The canonical workflow is four lines.

```r
library(climatekit)
frost_days <- ck_frost_days(tmin, dates, period = "annual")
spi3       <- ck_spi(precip, dates, scale = 3)
huglin     <- ck_huglin(tmin, tmax, dates, lat = 51.5)
```

Each call returns a tidy data frame that can be bound directly
into a panel with other indices or plotted without reshaping.

# Limitations

Five limitations apply.

1. \CRANpkg{climatekit} computes indices from daily observations.
   Sub-daily (hourly) indices, including heat-wave indices based
   on night-time minimum temperatures and apparent-temperature
   exceedances, are not implemented.
2. Gridded-climate input is not natively supported. Users must
   reduce a gridded product to a per-cell time series before
   passing it in. For CMIP6 and similar model output,
   \CRANpkg{terra}-based preprocessing is the natural partner.
3. The SPI and SPEI implementations use the gamma distribution as
   the parametric family. Users studying very arid regions may
   prefer the Pearson III family favoured by
   @vicente2010multiscalar; this is not currently exposed.
4. Quality control is the user's responsibility. Missing-value
   handling propagates `NA` through the aggregation functions.
5. Station homogenisation is not implemented. Detecting
   inhomogeneities (relocation, instrument change) in a daily
   record requires specialised techniques that are out of scope.

# Appendix of index definitions

Selected formulas used by \CRANpkg{climatekit}. Daily observations
are indexed by $t$; aggregations run over a reporting window
$\mathcal{T}$.

**Frost days.** $\mathrm{FD} = \sum_{t \in \mathcal{T}} \mathbf{1}\{T_{\min,t} < 0^\circ\mathrm{C}\}$.

**Growing degree days (base $b$).**
$\mathrm{GDD}_b = \sum_{t \in \mathcal{T}} \max(T_{\mathrm{mean},t} - b, 0)$.

**Huglin heliothermal index.** For latitude $\phi$ and
northern-hemisphere summation over 1 April to 30 September,
$$\mathrm{HI} = K(\phi) \sum_{t = \mathrm{Apr}}^{\mathrm{Sep}}
\frac{(T_{\mathrm{mean},t} - 10) + (T_{\max,t} - 10)}{2},$$
where $K(\phi)$ is Huglin's day-length correction.

**Standardised Precipitation Index.** For precipitation $P_s$
summed over an $s$-month window, fit a two-parameter gamma
distribution with shape $\alpha$ and scale $\beta$, compute the
non-exceedance probability $F(P_s; \alpha, \beta)$, and transform
to a standard normal: $\mathrm{SPI}_s = \Phi^{-1}(F(P_s; \alpha, \beta))$.

# Conclusion

Climate-index computation is a daily activity in meteorological
services, climate research, agricultural consultancies, and
insurance underwriting, and the R ecosystem had not had a single
unified implementation. \CRANpkg{climatekit} consolidates thirty
indices across temperature, precipitation, drought, agroclimatology,
and thermal comfort behind a tidy-data-frame output format. The
package is pure R, has no data dependencies, and returns output
that composes directly into downstream analysis.

# Acknowledgements {-}

The CCl/WCRP/JCOMM Expert Team on Climate Change Detection and
Indices (ETCCDI) and its successor the Expert Team on
Sector-specific Climate Indices (ET-SCI) developed and documented
the index definitions on which this package is built. The National
Oceanic and Atmospheric Administration publishes the Global
Historical Climatology Network Daily (GHCND) dataset used in the
case study. The author thanks the R Core Team and CRAN
maintainers for the infrastructure that supports open-source
statistical software.
